---
title: "Assignment 08"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}

## Homework 8

#### Due Friday, November 19, 11:59 PM CT

Problems 1-3 are based on the random variable $X \sim \text{Binomial}(60, 0.4)$.

### Problem 1

### A

Calculate and report the mean and standard deviation of the distirbution.

```{r}
n = 60
p = 0.4
mu = n*p
print(mu)
```

### B

Calculate and report $\prob(X = 25)$

```{r}
dbinom(25, n, p)
```

### C

Calculate and report $\prob(X >= 25)$

```{r}
1 - pbinom(25, n, p)
```

### D

Create a graph which displays the binomial distribution with bars for $x \ge 25$ colored in a different color from the rest to help visualize your answer to **C**.

```{r}
gbinom(n, p) +
  geom_binom_density(n, p, a = 25, color = "firebrick")
```

### Problem 2

What is the probability that $X$ is within one, two, and three standard deviations of the mean? Round each probability to four decimal places.

```{r}
sd = sqrt(n * p * (1 - p))
round(2 * (0.5 - pbinom(mu - sd, n, p)), digits = 4)
round(2 * (0.5 - pbinom(mu - 2 * sd, n, p)), digits = 4)
round(2 * (0.5 - pbinom(mu - 3 * sd, n, p)), digits = 4)
```


### Problem 3

Draw a graph of this binomial distribution with vertical lines with different line types at the end points of the intervals 1, 2, and 3 sigma from the mean.

- solid for $\mu \pm \sigma$
- dashed for $\mu \pm 2\sigma$
- dotted for $\mu \pm 3\sigma$

```{r}
gbinom(n, p) +
  geom_vline(xintercept = mu - sd) +
  geom_vline(xintercept = mu + sd) +
  geom_vline(xintercept = mu - 2 * sd, linetype = "dashed") +
  geom_vline(xintercept = mu + 2 * sd, linetype = "dashed") +
  geom_vline(xintercept = mu - 3 * sd, linetype = "dotted") +
  geom_vline(xintercept = mu + 3 * sd, linetype = "dotted") +
  labs(title = "Standard deviations of Binomial(60, 0.4) distribution", x = "outcome", y = "probability")
```



### Problem 4

### A

Create a tibble where each row corresponds to a binomial distribution with the following columns:

- `n` for $n = 1000,1001, \ldots, 10,000$
- `p` equal to 0.9 for each row
- `mu` equal to the mean
- `sigma` equal to the standard deviation
- `q.975` equal to the 0.975 quantile of the distribution
- `z = (q.975 - mu)/sigma`, the standardized z-score of the quantile
- `prob` equal to $\prob(X \ge q.975)$, a right-tail probability for the distribution in each row

Display all columns and the rows of this table for which $n$ is a multiple of 1000.

```{r}
bin_tbl = seq(1000, 10000) %>% 
  tbl_df() %>%
  rename(n = value) %>%
  mutate(p = 0.9, mu = n * p, sigma = sqrt(n * p * (1 - p)), q.975 = qbinom(0.975, n, p), z = (q.975 - mu) / sigma, prob = 1 - pbinom(q.975, n, p))

bin_tbl %>%
  filter(n %% 1000 == 0) %>%
  print(n = Inf, width = Inf)
```

### B

Use a line plot to show the relationship between $n$ on the x axis and $z$ on the y axis.
Add a smooth curve to highlight the trend.
Describe the pattern you see.

```{r}
bin_tbl %>% 
  ggplot(aes(x = n, y = z)) +
  geom_line() +
  geom_smooth() +
  labs(title = "Z vs n for binomial distribution with p = 0.9", x = "n", y = "Z")
```

$z$ approaches $1.95$ as $n$ increases.

### C

What is the average of the values of $z$ for $n > 9000$?

```{r}
bin_tbl %>% 
  filter(n > 9000) %>%
  pull("z") %>%
  mean()
```

### D

Plot the right tail probability `prob` on y axis versus `n` on the x axis.

```{r}
bin_tbl %>% 
  ggplot(aes(x = n, y = prob)) +
  geom_line() +
  labs(title = "Right tail probability vs n for binomial distribution", x = "n", y= "right tail probability")
```

### E

What is the average of the values of `prob` for $n > 9000$?

```{r}
bin_tbl %>% 
  filter(n > 9000) %>%
  pull("prob") %>%
  mean()
```


### Problem 5

### A

What is the 0.975 quantile of the standard normal distribution?

```{r}
qnorm(0.975)
```

### B

What is the area to the right of this quantile?

```{r}
1 - 0.975
```

### Problem 6

For the $\text{N}(500,100)$ distribution, find values $a$ and $b$ such that $a < b$, the area between $a$ and $b$ under the normal density is equal to 0.6,
and $a$ and $b$ are equally distant from $\mu = 500$.

```{r}
a = qnorm(0.2, 500, 100)
b = qnorm(0.8, 500, 100)
print(a)
print(b)
```


### Problem 7

Graph the $\text{N}(500,100)$ distribution and shade the area between $a$ and $b$ from the previous problem.

```{r}
gnorm(500, 100) +
  geom_norm_fill(500, 100, a = a, b = b) +
  labs(title = "Normal distribution center 60% quantile for N(500, 100)", x = "Value", y = "Density")
```





